---
title: "MovieLens Rating Prediction Project Report"
author: "Noam Kadosh"
date: "June 16th, 2020"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This project is part of the HarvardX course PH125.9x Data Science: Capstone project.
With the increasing popularity of streaming services, online news sites, social media, and the internet overall, recommendation systems become part of every service, website, or blog. The goal of these systems is to offer the user more content that will appeal to him and keep him on the website consuming more content. One application of machine learning is to predict and make recommendations to the user. The user will potentially like these recommendations, navigate to the relevant content, and consume more of it.
In 2006, Netflix offered a challenge to the data science community. Improve their recommendation system by at least 10% and win 1 million dollars. The goal was to predict the rating a specific user will give a particular movie. Then, the systems can recommend movies that users will rate high. We can understand from the big price money how important Netflix's recommendation system to the service is.

## Project Introduction

The project's goal is to use this dataset to predict the score a user will give a particular movie. There are various biases to consider when facing this problem. These can be social, geographical, cultural, psychological, and more. Each one of these can change the likings of every user.
I will train four different machine learning algorithms using the dataset. Starting with the most simple prediction algorithm, which is just the mean of the ratings, then considering different effects and combining them with other biases.
The algorithms will use the Root Mean Squared Error (RMSE) to evaluate performance. It is a way to measure the difference between the value observed to the value predicted, and the goal is to get it as low as possible.
The RMSE formula: 
$$ RMSE = \sqrt{\frac{1}{N}\displaystyle\sum_{u,i} (\hat{y}_{u,i}-y_{u,i})^{2}} $$
The target RMSE for this project is lower than 0.86490. 


## The Data
The data used for the project is the 10M version of MovieLens dataset, collected by GroupLens research lab at the University of Minnesota. The data contains movies, users, ratings, genres, and times.
The following code is included in the HarvardX capstone project course. This code splits the data to a training set and a validation set. The training set, edx in the code, is used to train the algorithms. The validation set, validation in the code, is used to test the algorithms on new never seen data ("the real world").

```{r data-preperation, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
##################################
# Create edx set, validation set #
##################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

 dl <- tempfile()
 download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
 ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                  col.names = c("userId", "movieId", "rating", "timestamp"))
 
 movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)

colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding")
# if using R 3.5 or earlier, use `set.seed(1)` instead
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

```{r additional-settings, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
# Additional libraries needed #
library(tidyr)
library(stringr)

options(digits = 5, pillar.sigfig = 5)
```

\pagebreak

# Methods and Analysis

## Data Analysis

Looking at the first few rows of the "edx", we can see the features which are "userId", "movieId", "rating", "timestamp", "title", and "genres". Each row represents a single rating a unique user gave a particular movie. 

```{r data-head, echo = FALSE}
head(edx)
```

A summary of the data can confirm that there are no missing values.

```{r data-summary, echo = FALSE}
summary(edx)
```

The total number of unique users is about 70,000. The number of unique movies is about 10,700.

```{r unique-values, echo = FALSE}
edx %>% summarize(num_users = n_distinct(userId), num_movies = n_distinct(movieId))
```

Ratings are on a scale of 0.5 to 5 with increments of 0.5. Users tend to rate movies relatively high. The most common rating is a rating of 4, followed by a rating of 3. 0.5 is the least common rating. We can infer from the chart that users tend to use whole numbers to rate movies.

```{r ratings-distribution, echo = TRUE}
# Ratings distribution
edx %>%
  ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.25, color = "black", fill = "#2e4057") +
  xlab("Rating") +
  scale_y_continuous(breaks = c(seq(0, 3000000, 500000))) +
  ylab("Count") +
  ggtitle("Rating distribution")
```

Users watch some movies more than others, for example, blockbuster films. Therefore they are being rated a lot more and usually higher than others. On the other hand, users rate some movies very few times, and some movies only once. These facts can result in untrustworthy estimations, especially for movies that are rated a few times. We will introduce a movie bias penalty term. This term controls the function used for prediction, which is very fluctuating, such that the coefficients don't take extreme values. Nonetheless, forecasts for movies rated only once will be challenging.

```{r movie-bias, echo = TRUE}
# Number of ratings per Movie - Movie bias
edx %>%
  count(movieId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 35, color = "black", fill = "#2e4057") +
  xlab("Ratings Count") +
  scale_x_log10() +
  ylab("Movie Count") +
  ggtitle("Number of Ratings per Movie")
```

Some users rate more movies than others, as seen in the first chart. The majority of users rate between 30 and 100 films. From the second chart, we can deduce that some users rate movies higher than others and some lower than others, possibly because some people are more positive and others negative. Another reason for that is that some users tend to rate movies they liked, thus rating on average higher, and some tend to rate movies they don't like, thus rating on average lower. We will introduce a user penalty term since some users don't rate a lot of movies.

```{r user-bias, echo = TRUE}
# Number of ratings given by users
edx %>%
  count(userId) %>%
  ggplot(aes(n)) +
  geom_histogram(bins = 35, color = "black", fill = "#2e4057") +
  xlab("Ratings Count") +
  scale_x_log10() +
  ylab("User Count") +
  ggtitle("Number of Ratings given by Users")
```

```{r mean-rating-by-user, echo = TRUE}
# Mean ratings given by user
edx %>%
  group_by(userId) %>%
  filter(n() >= 30) %>%
  summarize(mean_rating = mean(rating)) %>%
  ggplot(mean_rating) +
  geom_histogram(bins = 35,  color = "black", fill = "#2e4057") +
  xlab("Mean Rating") +
  ylab("Number of Users") +
  ggtitle("Mean Ratings Given By Users")
```

## Modeling






# Appendix

## Environment

Operating System:
```{r environment, echo=FALSE}
version
```
